---
title: "PHTO_DGE_Analysis"
author: "Harsha Tamtam"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
#this chuck is to setup defaults for basic settings of the webpage, note that this chunk is R language and the chunk below is CSS language
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	dev = c("png","pdf")
)
```

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```


## Introduction

## Library Section
```{r Libraries}
library(tidyverse)
library(dplyr)
library(tibble)
library(scales)
library(cowplot)
library(edgeR)
library(smplot2)
library(ggpubr)
library(workflowr)
library(ComplexHeatmap)
library(RColorBrewer)
library(gridExtra)
library(ggrepel)
library(ggVennDiagram)
library(grid)
library(ggrepel)
library(org.Hs.eg.db)
library(enrichplot)
library(clusterProfiler)
library(circlize)
library(gprofiler2)
library(stringr)
library(readr)
library(GO.db)
library(AnnotationDbi)
library(plotrix)
```

## File Calling
```{r}
PHTO_metadata <- readRDS("data/Emma_metadata_updated_all.RDS")
PHTO_counts <- read.csv("data/Emma_counts_file.csv")

working_counts_file <- PHTO_counts %>% 
  as.data.frame() %>% 
  rename(., Ensembl = gene_id) %>% 
  pivot_longer(-Ensembl, names_to = "Samples", values_to = "raw_counts") %>% 
  left_join(PHTO_metadata, by = c("Samples" = "Sample_ID")) %>% 
  dplyr::filter(!Stimulus %in% c("EtOH", "PFOA", "BPA")) %>%
  dplyr::filter(Time == 24) %>% 
  dplyr::select(Ensembl, raw_counts, Final_sample_name) %>% 
  pivot_wider(names_from = Final_sample_name, values_from = raw_counts) %>% 
  column_to_rownames("Ensembl")

working_metadata_file <- PHTO_metadata %>% 
  dplyr::filter(!Stimulus %in% c("EtOH", "PFOA", "BPA")) %>%
  dplyr::filter(Time == 24) %>% 
  dplyr::mutate(
    Grouping = case_when(
      Species == "C" & Stimulus == "DMSO" ~ "Chimp_DMSO",
      Species == "H" & Stimulus == "DMSO" ~ "Human_DMSO",
      Species == "C" & Stimulus == "H2O" ~ "Chimp_H2O",
      Species == "H" & Stimulus == "H2O" ~ "Human_H2O",
      Species == "C" & Stimulus == "DOX" ~ "Chimp_DOX",
      Species == "H" & Stimulus == "DOX" ~ "Human_DOX",
      Species == "C" & Stimulus == "NUTL" ~ "Chimp_NUTL",
      Species == "H" & Stimulus == "NUTL" ~ "Human_NUTL",
      Species == "C" & Stimulus == "TUN" ~ "Chimp_TUN",
      Species == "H" & Stimulus == "TUN" ~ "Human_TUN",
      Species == "C" & Stimulus == "THA" ~ "Chimp_THA",
      Species == "H" & Stimulus == "THA" ~ "Human_THA",
      Species == "C" & Stimulus == "LPS" ~ "Chimp_LPS",
      Species == "H" & Stimulus == "LPS" ~ "Human_LPS",
      Species == "C" & Stimulus == "TNFa" ~ "Chimp_TNFa",
      Species == "H" & Stimulus == "TNFa" ~ "Human_TNFa"
  )
)

```

## Converting to Log2 CPM and filtering rowMeans log2cpm > 0:
```{r}
log2cpm <- cpm(working_counts_file, log = TRUE)
filtered_counts <- working_counts_file[rowMeans(log2cpm)> 0,]

# dim(filtered_counts)
# str(filtered_counts)
```

I will be ensuring my factors for the contrast matrices I will make are set correctly, but I need to make new data tables for each treatment type

## Identifying DGEs for each Response Set (DDR, UPR, IMR)

### DNA Damage Response (DDR) DGE, Voom, and Contrast Matrix Development

#### Using model matrix (~ Species * Stimulus)
```{r eval=FALSE, include=FALSE}
# ----------------------------------------------------------------------------------------------------------------------------------------------------
# I will be ensuring my factors for the contrast matrices I will make are set correctly, but I need to make new data tables for each treatment type
# ----------------------------------------------------------------------------------------------------------------------------------------------------

# # DDR setup
DDR_metadata <- working_metadata_file %>%
  dplyr::filter(Stimulus %in% c("DMSO", "DOX", "NUTL"))

DDR_counts <- filtered_counts %>%
  dplyr::select(all_of(DDR_metadata$Final_sample_name))

ncol(DDR_counts) # Run this with the line below to ensure that sample numbers match
nrow(DDR_metadata) # Run this with the line above to ensure that sample numbers match

DDR_counts <- DDR_counts[, DDR_metadata$Final_sample_name] # Run this to explicitly ensure sample ordering is identical between counts and metadata file

# # Creating DDR-specific DGE list
DDR_dge <- DGEList(
  counts = DDR_counts
)

DDR_dge <- calcNormFactors(
  DDR_dge,
  method = "TMM"
)

# # Building the DDR design matrix
DDR_metadata$Stimulus <- factor(
  DDR_metadata$Stimulus,
  levels = c("DMSO", "DOX", "NUTL")
)

DDR_metadata$Species <- relevel(
  factor(DDR_metadata$Species),
  ref = "C"
)

DDR_design <- model.matrix(
  ~ Species * Stimulus,
  data = DDR_metadata
)

# colnames(DDR_design)

# Renaming columns to use underscores instead of default colons, colons will break the code
colnames(DDR_design) <- c(
  "Intercept",
  "SpeciesH",
  "StimulusDOX",                      
  "StimulusNUTL",
  "SpeciesH_StimulusDOX",
  "SpeciesH_StimulusNUTL"
)

# Running Duplicate Correlations
DDR_cor_fit <- duplicateCorrelation(object = DDR_dge$counts,
                                design = DDR_design,
                                block = DDR_metadata$Ind
                                )

DDR_voom_trend <- voom(DDR_dge,
                   DDR_design,
                   block = DDR_metadata$Ind, 
                   correlation = DDR_cor_fit$consensus.correlation,
                   plot = FALSE)


# Fitting linear model
DDR_fit <- lmFit(
  DDR_voom_trend,
  DDR_design,
  block = DDR_metadata$Ind,
  correlation = DDR_cor_fit$consensus.correlation
)

head(design)

# Defining contrasts
DDR_contrast_matrix <- makeContrasts(
  Human_vs_Chimp_DMSO = SpeciesH,
  Human_vs_Chimp_DOX = SpeciesH + SpeciesH_StimulusDOX,
  Human_vs_Chimp_NUTL = SpeciesH + SpeciesH_StimulusNUTL,
  Chimp_DOX_vs_DMSO = StimulusDOX,
  Human_DOX_vs_DMSO = StimulusDOX + SpeciesH_StimulusDOX,
  Chimp_NUTL_vs_DMSO = StimulusNUTL,
  Human_NUTL_vs_DMSO = StimulusNUTL + SpeciesH_StimulusNUTL,
  Interaction_DOX = SpeciesH_StimulusDOX,
  Interaction_NUTL = SpeciesH_StimulusNUTL,
  levels = DDR_design
)

# Applying contrasts across samples
DDR_fit2 <- contrasts.fit(DDR_fit, DDR_contrast_matrix)
DDR_efit2 <- eBayes(DDR_fit2)
DDR_signif_gene_results = decideTests(DDR_efit2, adjust.method = "BH", p.value = 0.05)

summary(DDR_signif_gene_results)
plotSA(DDR_efit2, main = "Mean-variance DDR final model trend")

```

#### Using model matrix (~ 0 + Grouping)
```{r eval=FALSE, include=FALSE}
# ----------------------------------------------------------------------------------------------------------------------------------------------------
# I will be ensuring my factors for the contrast matrices I will make are set correctly, but I need to make new data tables for each treatment type
# ----------------------------------------------------------------------------------------------------------------------------------------------------

# # DDR setup
DDR_metadata <- working_metadata_file %>%
  dplyr::filter(Stimulus %in% c("DMSO", "DOX", "NUTL"))

DDR_counts <- filtered_counts %>%
  dplyr::select(all_of(DDR_metadata$Final_sample_name))

ncol(DDR_counts) # Run this with the line below to ensure that sample numbers match
nrow(DDR_metadata) # Run this with the line above to ensure that sample numbers match

DDR_counts <- DDR_counts[, DDR_metadata$Final_sample_name] # Run this to explicitly ensure sample ordering is identical between counts and metadata file

# # Creating DDR-specific DGE list
DDR_dge <- DGEList(
  counts = DDR_counts
)

DDR_dge <- calcNormFactors(
  DDR_dge,
  method = "TMM"
)

# # Building the DDR design matrix
DDR_metadata$Stimulus <- factor(
  DDR_metadata$Stimulus,
  levels = c("DMSO", "DOX", "NUTL")
)

DDR_metadata$Species <- relevel(
  factor(DDR_metadata$Species),
  ref = "C"
)

DDR_design <- model.matrix(
  ~ Species + Stimulus + Species:Stimulus,
  data = DDR_metadata
)

colnames(DDR_design)
```



#### Using model matrix (~ 0 + Species:Stimulus)
```{r eval=FALSE, include=FALSE}
# ----------------------------------------------------------------------------------------------------------------------------------------------------
# I will be ensuring my factors for the contrast matrices I will make are set correctly, but I need to make new data tables for each treatment type
# ----------------------------------------------------------------------------------------------------------------------------------------------------

# # DDR setup
DDR_metadata <- working_metadata_file %>%
  dplyr::filter(Stimulus %in% c("DMSO", "DOX", "NUTL"))

DDR_counts <- filtered_counts %>%
  dplyr::select(all_of(DDR_metadata$Final_sample_name))

ncol(DDR_counts) # Run this with the line below to ensure that sample numbers match
nrow(DDR_metadata) # Run this with the line above to ensure that sample numbers match

DDR_counts <- DDR_counts[, DDR_metadata$Final_sample_name] # Run this to explicitly ensure sample ordering is identical between counts and metadata file

# # Creating DDR-specific DGE list
DDR_dge <- DGEList(
  counts = DDR_counts
)

DDR_dge <- calcNormFactors(
  DDR_dge,
  method = "TMM"
)

# # Building the DDR design matrix
DDR_metadata$Stimulus <- factor(
  DDR_metadata$Stimulus,
  levels = c("DMSO", "DOX", "NUTL")
)

DDR_metadata$Species <- relevel(
  factor(DDR_metadata$Species),
  ref = "C"
)

DDR_design <- model.matrix(
  ~ 0 + Species:Stimulus,
  data = DDR_metadata
)

colnames(DDR_design)

# Renaming columns to use underscores instead of default colons, colons will break the code
colnames(DDR_design) <- c(
  "SpeciesC_StimulusDMSO",
  "SpeciesH_StimulusDMSO",
  "SpeciesC_StimulusDOX",                      
  "SpeciesH_StimulusDOX",
  "SpeciesC_StimulusNUTL",
  "SpeciesH_StimulusNUTL"
)

# Running Duplicate Correlations
DDR_cor_fit <- duplicateCorrelation(object = DDR_dge$counts,
                                design = DDR_design,
                                block = DDR_metadata$Ind
                                )

DDR_voom_trend <- voom(DDR_dge,
                   DDR_design,
                   block = DDR_metadata$Ind, 
                   correlation = DDR_cor_fit$consensus.correlation,
                   plot = FALSE)


# Fitting linear model
DDR_fit <- lmFit(
  DDR_voom_trend,
  DDR_design,
  block = DDR_metadata$Ind,
  correlation = DDR_cor_fit$consensus.correlation
)

head(design)

# Defining contrasts
DDR_contrast_matrix <- makeContrasts(
  Human_vs_Chimp_DMSO = 
    (SpeciesH_StimulusDMSO - SpeciesC_StimulusDMSO),
  Human_vs_Chimp_DOX = 
    (SpeciesH_StimulusDOX - SpeciesC_StimulusDOX),
  Human_vs_Chimp_NUTL = 
    (SpeciesH_StimulusNUTL - SpeciesC_StimulusNUTL),
  Chimp_DOX_vs_DMSO = 
    (SpeciesC_StimulusDOX - SpeciesC_StimulusDMSO),
  Human_DOX_vs_DMSO = 
    (SpeciesH_StimulusDOX - SpeciesH_StimulusDMSO),
  Chimp_NUTL_vs_DMSO = 
    (SpeciesC_StimulusNUTL - SpeciesC_StimulusDMSO),
  Human_NUTL_vs_DMSO = 
    (SpeciesH_StimulusNUTL - SpeciesH_StimulusDMSO),
  Interaction_DOX = 
    (SpeciesH_StimulusDOX - SpeciesH_StimulusDMSO) -
    (SpeciesC_StimulusDOX - SpeciesC_StimulusDMSO),
  Interaction_NUTL = 
    (SpeciesH_StimulusNUTL - SpeciesH_StimulusDMSO) -
    (SpeciesC_StimulusNUTL - SpeciesC_StimulusDMSO),
  levels = DDR_design
)

# Applying contrasts across samples
DDR_fit2 <- contrasts.fit(DDR_fit, DDR_contrast_matrix)
DDR_efit2 <- eBayes(DDR_fit2)
DDR_signif_gene_results = decideTests(DDR_efit2, adjust.method = "BH", p.value = 0.05)

summary(DDR_signif_gene_results)
plotSA(DDR_efit2, main = "Mean-variance DDR final model trend")
```

#### Comparison of model.matricies for DDR
```{r eval=FALSE, include=FALSE}
design1 <- model.matrix(
  ~ Species * Stimulus,
  data = DDR_metadata
)

# colnames(design1)

colnames(design1) <- c(
  "Intercept",
  "SpeciesH",
  "StimulusDOX",                      
  "StimulusNUTL",
  "SpeciesH_StimulusDOX",
  "SpeciesH_StimulusNUTL"
)

contrast1 <- makeContrasts(
  Human_DOX_vs_VEH =
    StimulusDOX + SpeciesH_StimulusDOX,
  levels = design1
)

first_cor_fit <- duplicateCorrelation(object = DDR_dge$counts,
                                design = design1,
                                block = DDR_metadata$Ind
                                )

first_voom_trend <- voom(DDR_dge,
                   design1,
                   block = DDR_metadata$Ind, 
                   correlation = first_cor_fit$consensus.correlation,
                   plot = TRUE)


# Fitting linear model
first_fit <- lmFit(
  first_voom_trend,
  design1,
  block = DDR_metadata$Ind,
  correlation = first_cor_fit$consensus.correlation
)

  # Applying contrasts across samples
first_fit2 <- contrasts.fit(first_fit, contrast1)
first_efit2 <- eBayes(first_fit2)
first_signif_gene_results = decideTests(first_efit2, adjust.method = "BH", p.value = 0.05)

summary(first_signif_gene_results)
plotSA(first_efit2, main = "Mean-variance First final model trend")

# ---------------------------------------------------------------------------------------------------------------------------
design2 <- model.matrix(
  ~ 0 + Species:Stimulus,
  data = DDR_metadata
)

# colnames(design2)

colnames(design2) <- c(
  "SpeciesC_StimulusDMSO",
  "SpeciesH_StimulusDMSO",
  "SpeciesC_StimulusDOX",                      
  "SpeciesH_StimulusDOX",
  "SpeciesC_StimulusNUTL",
  "SpeciesH_StimulusNUTL"
)

contrast2 <- makeContrasts(
  Human_DOX_vs_VEH =
    (SpeciesH_StimulusDOX) -
    (SpeciesH_StimulusDMSO),
  levels = design2
)

second_cor_fit <- duplicateCorrelation(object = DDR_dge$counts,
                                design = design2,
                                block = DDR_metadata$Ind
                                )

second_voom_trend <- voom(DDR_dge,
                   design2,
                   block = DDR_metadata$Ind, 
                   correlation = second_cor_fit$consensus.correlation,
                   plot = TRUE)


# Fitting linear model
second_fit <- lmFit(
  second_voom_trend,
  design2,
  block = DDR_metadata$Ind,
  correlation = second_cor_fit$consensus.correlation
)

  # Applying contrasts across samples
second_fit2 <- contrasts.fit(second_fit, contrast1)
second_efit2 <- eBayes(second_fit2)
second_signif_gene_results = decideTests(second_efit2, adjust.method = "BH", p.value = 0.05)

summary(second_signif_gene_results)
plotSA(second_efit2, main = "Mean-variance Second final model trend")

results1 <- topTable(
  first_efit2,
  coef = "Human_DOX_vs_VEH",
  number = Inf,
  sort.by = "none"
)

results2 <- topTable(
  second_efit2,
  coef = "Human_DOX_vs_VEH",
  number = Inf,
  sort.by = "none"
)

cor(
  results1$logFC,
  results2$logFC
)
```


### Unfolding Protein Response (UPR) DGE, Voom, and Contrast Matrix Development
```{r eval=FALSE, include=FALSE}
# ----------------------------------------------------------------------------------------------------------------------------------------------------
# I will be ensuring my factors for the contrast matrices I will make are set correctly, but I need to make new data tables for each treatment type
# ----------------------------------------------------------------------------------------------------------------------------------------------------

# # UPR setup
UPR_metadata <- working_metadata_file %>%
  dplyr::filter(Stimulus %in% c("DMSO", "THA", "TUN"))

UPR_counts <- filtered_counts %>%
  dplyr::select(all_of(UPR_metadata$Final_sample_name))

ncol(UPR_counts) # Run this with the line below to ensure that sample numbers match
nrow(UPR_metadata) # Run this with the line above to ensure that sample numbers match

UPR_counts <- UPR_counts[, UPR_metadata$Final_sample_name] # Run this to explicitly ensure sample ordering is identical between counts and metadata file

# # Creating DDR-specific DGE list
UPR_dge <- DGEList(
  counts = UPR_counts
)

UPR_dge <- calcNormFactors(
  UPR_dge,
  method = "TMM"
)

# # Building the DDR design matrix
UPR_metadata$Stimulus <- factor(
  UPR_metadata$Stimulus,
  levels = c("DMSO", "THA", "TUN")
)

UPR_metadata$Species <- relevel(
  factor(UPR_metadata$Species),
  ref = "C"
)

UPR_design <- model.matrix(
  ~ Species * Stimulus,
  data = UPR_metadata
)

colnames(UPR_design)

# Renaming columns to use underscores instead of default colons, colons will break the code
colnames(UPR_design) <- c(
  "Intercept",
  "SpeciesH",
  "StimulusTHA",                      
  "StimulusTUN",
  "SpeciesH_StimulusTHA",
  "SpeciesH_StimulusTUN"
)

# Running Duplicate Correlations
UPR_cor_fit <- duplicateCorrelation(object = UPR_dge$counts,
                                design = UPR_design,
                                block = UPR_metadata$Ind
                                )

UPR_voom_trend <- voom(UPR_dge,
                   UPR_design,
                   block = UPR_metadata$Ind, 
                   correlation = UPR_cor_fit$consensus.correlation,
                   plot = TRUE)


# Fitting linear model
UPR_fit <- lmFit(
  UPR_voom_trend,
  UPR_design,
  block = UPR_metadata$Ind,
  correlation = UPR_cor_fit$consensus.correlation
)

head(UPR_design)

# Defining contrasts
UPR_contrast_matrix <- makeContrasts(
  Human_vs_Chimp_DMSO = SpeciesH,
  Human_vs_Chimp_THA = SpeciesH + SpeciesH_StimulusTHA,
  Human_vs_Chimp_TUN = SpeciesH + SpeciesH_StimulusTUN,
  Chimp_THA_vs_DMSO = StimulusTHA,
  Human_THA_vs_DMSO = StimulusTHA + SpeciesH_StimulusTHA,
  Chimp_TUN_vs_DMSO = StimulusTUN,
  Human_TUN_vs_DMSO = StimulusTUN + SpeciesH_StimulusTUN,
  Interaction_THA = SpeciesH_StimulusTHA,
  Interaction_TUN = SpeciesH_StimulusTUN,
  levels = UPR_design
)

# Applying contrasts across samples
UPR_fit2 <- contrasts.fit(UPR_fit, UPR_contrast_matrix)
UPR_efit2 <- eBayes(UPR_fit2)
UPR_signif_gene_results = decideTests(UPR_efit2, adjust.method = "BH", p.value = 0.05)

summary(UPR_signif_gene_results)
plotSA(UPR_efit2, main = "Mean-variance UPR final model trend")

```


### Innate Immune Response (IMR) DGE, Voom, and Contrast Matrix Development
```{r eval=FALSE, include=FALSE}
# ----------------------------------------------------------------------------------------------------------------------------------------------------
# I will be ensuring my factors for the contrast matrices I will make are set correctly, but I need to make new data tables for each treatment type
# ----------------------------------------------------------------------------------------------------------------------------------------------------

# # IMR setup
IMR_metadata <- working_metadata_file %>%
  dplyr::filter(Stimulus %in% c("H2O", "LPS", "TNFa"))

IMR_counts <- filtered_counts %>%
  dplyr::select(all_of(IMR_metadata$Final_sample_name))

ncol(IMR_counts) # Run this with the line below to ensure that sample numbers match
nrow(IMR_metadata) # Run this with the line above to ensure that sample numbers match

IMR_counts <- IMR_counts[, IMR_metadata$Final_sample_name] # Run this to explicitly ensure sample ordering is identical between counts and metadata file

# # Creating DDR-specific DGE list
IMR_dge <- DGEList(
  counts = IMR_counts
)

IMR_dge <- calcNormFactors(
  IMR_dge,
  method = "TMM"
)

# # Building the DDR design matrix
IMR_metadata$Stimulus <- factor(
  IMR_metadata$Stimulus,
  levels = c("H2O", "LPS", "TNFa")
)

IMR_metadata$Species <- relevel(
  factor(IMR_metadata$Species),
  ref = "C"
)

IMR_design <- model.matrix(
  ~ Species * Stimulus,
  data = IMR_metadata
)

colnames(IMR_design)

# Renaming columns to use underscores instead of default colons, colons will break the code
colnames(IMR_design) <- c(
  "Intercept",
  "SpeciesH",
  "StimulusLPS",                      
  "StimulusTNFa",
  "SpeciesH_StimulusLPS",
  "SpeciesH_StimulusTNFa"
)

# Running Duplicate Correlations
IMR_cor_fit <- duplicateCorrelation(object = IMR_dge$counts,
                                design = IMR_design,
                                block = IMR_metadata$Ind
                                )

IMR_voom_trend <- voom(IMR_dge,
                   IMR_design,
                   block = IMR_metadata$Ind, 
                   correlation = IMR_cor_fit$consensus.correlation,
                   plot = TRUE)


# Fitting linear model
IMR_fit <- lmFit(
  IMR_voom_trend,
  IMR_design,
  block = IMR_metadata$Ind,
  correlation = IMR_cor_fit$consensus.correlation
)

head(IMR_design)

# Defining contrasts
IMR_contrast_matrix <- makeContrasts(
  Human_vs_Chimp_H2O = SpeciesH,
  Human_vs_Chimp_LPS = SpeciesH + SpeciesH_StimulusLPS,
  Human_vs_Chimp_TNFa = SpeciesH + SpeciesH_StimulusTNFa,
  Chimp_LPS_vs_H2O = StimulusLPS,
  Human_LPS_vs_H2O = StimulusLPS + SpeciesH_StimulusLPS,
  Chimp_TNFa_vs_H2O = StimulusTNFa,
  Human_TNFa_vs_H2O = StimulusTNFa + SpeciesH_StimulusTNFa,
  Interaction_THA = SpeciesH_StimulusLPS,
  Interaction_TNFa = SpeciesH_StimulusTNFa,
  levels = IMR_design
)

# Applying contrasts across samples
IMR_fit2 <- contrasts.fit(IMR_fit, IMR_contrast_matrix)
IMR_efit2 <- eBayes(IMR_fit2)
IMR_signif_gene_results = decideTests(IMR_efit2, adjust.method = "BH", p.value = 0.05)

summary(IMR_signif_gene_results)
plotSA(IMR_efit2, main = "Mean-variance IMR final model trend")

```


## Identifying DGEs for ALL Responses as One Set

### Using model.matrix(~0 + Species:Stimulus)
```{r eval=FALSE, include=FALSE}

filtered_counts <- filtered_counts[, working_metadata_file$Final_sample_name] # Run this to explicitly ensure sample ordering is identical between counts and metadata file

# # Creating DDR-specific DGE list
all_dge <- DGEList(
  counts = filtered_counts
)

all_dge <- calcNormFactors(
  all_dge,
  method = "TMM"
)

# # Building the DDR design matrix
working_metadata_file$Stimulus <- factor(
  working_metadata_file$Stimulus,
  levels = c("DMSO",
             "DOX",
             "NUTL",
             "TUN",
             "THA",
             "H2O",
             "TNFa",
             "LPS"
             )
  )

working_metadata_file$Species <- relevel(
  factor(working_metadata_file$Species),
  ref = "C"
)

all_dge_design <- model.matrix(
  ~ 0 + Species:Stimulus,
  data = working_metadata_file
)

head(all_dge_design)
colnames(all_dge_design)

# Renaming columns to use underscores instead of default colons, colons will break the code
colnames(all_dge_design) <- c(
  "SpeciesC_StimulusDMSO",
  "SpeciesH_StimulusDMSO",
  "SpeciesC_StimulusDOX",
  "SpeciesH_StimulusDOX",                      
  "SpeciesC_StimulusNUTL",
  "SpeciesH_StimulusNUTL",
  "SpeciesC_StimulusTUN",
  "SpeciesH_StimulusTUN",
  "SpeciesC_StimulusTHA",
  "SpeciesH_StimulusTHA",
  "SpeciesC_StimulusH2O",
  "SpeciesH_StimulusH2O",
  "SpeciesC_StimulusTNFa",
  "SpeciesH_StimulusTNFa",
  "SpeciesC_StimulusLPS",
  "SpeciesH_StimulusLPS"
)

# Running Duplicate Correlations
all_dge_cor_fit <- duplicateCorrelation(object = all_dge$counts,
                                design = all_dge_design,
                                block = working_metadata_file$Ind
                                )

all_dge_voom_trend <- voom(all_dge,
                   all_dge_design,
                   block = working_metadata_file$Ind, 
                   correlation = all_dge_cor_fit$consensus.correlation,
                   plot = FALSE)


# Fitting linear model
all_dge_fit <- lmFit(
  all_dge_voom_trend,
  all_dge_design,
  block = working_metadata_file$Ind,
  correlation = all_dge_cor_fit$consensus.correlation
)

head(all_dge_design)

# Defining contrasts
all_dge_contrast_matrix <- makeContrasts(
  Human_DOX_vs_DMSO =
    SpeciesH_StimulusDOX - SpeciesH_StimulusDMSO,
  Chimp_DOX_vs_DMSO =
    SpeciesC_StimulusDOX - SpeciesC_StimulusDMSO,
  Interaction_DOX =
    (SpeciesH_StimulusDOX - SpeciesH_StimulusDMSO) -
    (SpeciesC_StimulusDOX - SpeciesC_StimulusDMSO),
  Human_NUTL_vs_DMSO =
    SpeciesH_StimulusNUTL - SpeciesH_StimulusDMSO,
  Chimp_NUTL_vs_DMSO =
    SpeciesC_StimulusNUTL - SpeciesC_StimulusDMSO,
  Interaction_NUTL =
    (SpeciesH_StimulusNUTL - SpeciesH_StimulusDMSO) -
    (SpeciesC_StimulusNUTL - SpeciesC_StimulusDMSO),
  Human_TUN_vs_DMSO =
     SpeciesH_StimulusTUN - SpeciesH_StimulusDMSO,
  Chimp_TUN_vs_DMSO =
    SpeciesC_StimulusTUN - SpeciesC_StimulusDMSO,
  Interaction_TUN =
    (SpeciesH_StimulusTUN - SpeciesH_StimulusDMSO) -
    (SpeciesC_StimulusTUN - SpeciesC_StimulusDMSO),
  Human_THA_vs_DMSO =
    SpeciesH_StimulusTHA - SpeciesH_StimulusDMSO,
  Chimp_THA_vs_DMSO =
    SpeciesC_StimulusTHA - SpeciesC_StimulusDMSO,
  Interaction_THA =
    (SpeciesH_StimulusTHA - SpeciesH_StimulusDMSO) -
    (SpeciesC_StimulusTHA - SpeciesC_StimulusDMSO),
  Human_TNFa_vs_H2O =
    SpeciesH_StimulusTNFa - SpeciesH_StimulusH2O,
  Chimp_TNFa_vs_H2O =
    SpeciesC_StimulusTNFa - SpeciesC_StimulusH2O,
  Interaction_TNFa =
    (SpeciesH_StimulusTNFa - SpeciesH_StimulusH2O) -
    (SpeciesC_StimulusTNFa - SpeciesC_StimulusH2O),
  Human_LPS_vs_H2O =
    SpeciesH_StimulusLPS - SpeciesH_StimulusH2O,
  Chimp_LPS_vs_H2O =
   SpeciesC_StimulusLPS - SpeciesC_StimulusH2O,
  Interaction_LPS =
    (SpeciesH_StimulusLPS - SpeciesH_StimulusH2O) -
    (SpeciesC_StimulusLPS - SpeciesC_StimulusH2O),
  levels = all_dge_design
)

# Applying contrasts across samples
all_dge_fit2 <- contrasts.fit(all_dge_fit, all_dge_contrast_matrix)
all_dge_efit2 <- eBayes(all_dge_fit2)
all_dge_signif_gene_results = decideTests(all_dge_efit2, adjust.method = "BH", p.value = 0.05)

summary(all_dge_signif_gene_results)
plotSA(all_dge_efit2, main = "Mean-variance All Response Samples final model trend")
```

### Using model.matrix(~0 + Grouping)
```{r eval=FALSE, include=FALSE}
filtered_counts <- filtered_counts[, working_metadata_file$Final_sample_name] # Run this to explicitly ensure sample ordering is identical between counts and metadata file

# # Creating DDR-specific DGE list
all_dge <- DGEList(
  counts = filtered_counts
)

all_dge <- calcNormFactors(
  all_dge,
  method = "TMM"
)

# # Building the DDR design matrix
working_metadata_file$Stimulus <- factor(
  working_metadata_file$Stimulus,
  levels = c("DMSO",
             "DOX",
             "NUTL",
             "TUN",
             "THA",
             "H2O",
             "TNFa",
             "LPS"
             )
  )

# working_metadata_file$Species <- relevel(
#   factor(working_metadata_file$Species),
#   ref = "C"
# )

all_dge_design <- model.matrix(
  ~ 0 + Grouping,
  data = working_metadata_file
)

# head(all_dge_design)
colnames(all_dge_design)

# Running Duplicate Correlations
all_dge_cor_fit <- duplicateCorrelation(object = all_dge$counts,
                                design = all_dge_design,
                                block = working_metadata_file$Ind
                                )

all_dge_voom_trend <- voom(all_dge,
                   all_dge_design,
                   block = working_metadata_file$Ind,
                   correlation = all_dge_cor_fit$consensus.correlation,
                   plot = FALSE)


# Fitting linear model
all_dge_fit <- lmFit(
  all_dge_voom_trend,
  all_dge_design,
  block = working_metadata_file$Ind,
  correlation = all_dge_cor_fit$consensus.correlation
)

# Defining contrasts
all_dge_contrast_matrix <- makeContrasts(
  Human_DOX_vs_DMSO =
    GroupingHuman_DOX - GroupingHuman_DMSO,
  Chimp_DOX_vs_DMSO =
    GroupingChimp_DOX - GroupingChimp_DMSO,
  Interaction_DOX =
    (GroupingHuman_DOX - GroupingHuman_DMSO) -
    (GroupingChimp_DOX - GroupingChimp_DMSO),
  Human_NUTL_vs_DMSO =
    GroupingHuman_NUTL - GroupingHuman_DMSO,
  Chimp_NUTL_vs_DMSO =
    GroupingChimp_NUTL - GroupingChimp_DMSO,
  Interaction_NUTL =
    (GroupingHuman_NUTL - GroupingHuman_DMSO) -
    (GroupingChimp_NUTL - GroupingChimp_DMSO),
  Human_TUN_vs_DMSO =
     GroupingHuman_TUN - GroupingHuman_DMSO,
  Chimp_TUN_vs_DMSO =
    GroupingChimp_TUN - GroupingChimp_DMSO,
  Interaction_TUN =
    (GroupingHuman_TUN - GroupingHuman_DMSO) -
    (GroupingChimp_TUN - GroupingChimp_DMSO),
  Human_THA_vs_DMSO =
    GroupingHuman_THA - GroupingHuman_DMSO,
  Chimp_THA_vs_DMSO =
    GroupingChimp_THA - GroupingChimp_DMSO,
  Interaction_THA =
    (GroupingHuman_THA - GroupingHuman_DMSO) -
    (GroupingChimp_THA - GroupingChimp_DMSO),
  Human_TNFa_vs_H2O =
    GroupingHuman_TNFa - GroupingHuman_H2O,
  Chimp_TNFa_vs_H2O =
    GroupingChimp_TNFa - GroupingChimp_H2O,
  Interaction_TNFa =
    (GroupingHuman_TNFa - GroupingHuman_H2O) -
    (GroupingChimp_TNFa - GroupingChimp_H2O),
  Human_LPS_vs_H2O =
    GroupingHuman_LPS - GroupingHuman_H2O,
  Chimp_LPS_vs_H2O =
   GroupingChimp_LPS - GroupingChimp_H2O,
  Interaction_LPS =
    (GroupingHuman_LPS - GroupingHuman_H2O) -
    (GroupingChimp_LPS - GroupingChimp_H2O),
  levels = all_dge_design
)

# Applying contrasts across samples
all_dge_fit2 <- contrasts.fit(all_dge_fit, all_dge_contrast_matrix)
all_dge_efit2 <- eBayes(all_dge_fit2)
all_dge_signif_gene_results = decideTests(all_dge_efit2, adjust.method = "BH", p.value = 0.05)

summary(all_dge_signif_gene_results)
plotSA(all_dge_efit2, main = "Mean-variance All Response Samples final model trend")
```

### Using model.matrix (~ Species * Stimulus), using Chimp and H2O as my references
```{r}
filtered_counts <- filtered_counts[, working_metadata_file$Final_sample_name] # Run this to explicitly ensure sample ordering is identical between counts and metadata file

# # Creating DDR-specific DGE list
all_dge <- DGEList(
  counts = filtered_counts
)

all_dge <- calcNormFactors(
  all_dge,
  method = "TMM"
)

# # Building the DDR design matrix
working_metadata_file$Stimulus <- factor(
  working_metadata_file$Stimulus,
  levels = c("H2O", # Water is leveled first because this will be our reference
             "DMSO",
             "DOX",
             "NUTL",
             "TUN",
             "THA",
             "TNFa",
             "LPS"
             )
  )

working_metadata_file$Species <- relevel(
  factor(working_metadata_file$Species),
  ref = "C"
)

working_metadata_file$Stimulus <- relevel(
  factor(working_metadata_file$Stimulus),
  ref = "H2O"
) # We are setting the Stimulus reference to H2O because DMSO is dissolved in H2O

all_dge_design <- model.matrix(
  ~ Species * Stimulus, # ORDER MATTERS: here, with species first and stimulus second, I am controlling for species while observing interactions of the main effect, stimulus
  data = working_metadata_file
)

colnames(all_dge_design)
# all_dge_design

colnames(all_dge_design) <- c(
  "Intercept", # means Chimp + H2O
  "SpeciesH", # means Human vs Chimp under H2O
  "StimulusDMSO", # means Chimp DMSO vs H2O
  "StimulusDOX", # means Chimp DOX vs H2O
  "StimulusNUTL", # means Chimp NUTL vs H2O
  "StimulusTUN", # means Chimp TUN vs H2O
  "StimulusTHA", # means Chimp THA vs H2O
  "StimulusTNFa", # means Chimp TNFa vs H2O
  "StimulusLPS", # means Chimp LPS vs H2O
  "SpeciesH_StimulusDMSO", # means (Human DMSO vs H2O) - (Chimp DMSO vs H2O)
  "SpeciesH_StimulusDOX", # means (Human DOX vs H2O) - (Chimp DOX vs H2O)
  "SpeciesH_StimulusNUTL", # means (Human NUTL vs H2O) - (Chimp NUTL vs H2O)
  "SpeciesH_StimulusTUN", # means (Human TUN vs H2O) - (Chimp TUN vs H2O)
  "SpeciesH_StimulusTHA", # means (Human THA vs H2O) - (Chimp THA vs H2O)
  "SpeciesH_StimulusTNFa", # means (Human TNFa vs H2O) - (Chimp TNFa vs H2O)
  "SpeciesH_StimulusLPS" # means (Human LPS vs H2O) - (Chimp LPS vs H2O)
)

# Running Duplicate Correlations
all_dge_cor_fit <- duplicateCorrelation(object = all_dge$counts,
                                design = all_dge_design,
                                block = working_metadata_file$Ind
                                )

all_dge_voom_trend <- voom(all_dge,
                   all_dge_design,
                   block = working_metadata_file$Ind, 
                   correlation = all_dge_cor_fit$consensus.correlation,
                   plot = FALSE)


# Fitting linear model
all_dge_fit <- lmFit(
  all_dge_voom_trend,
  all_dge_design,
  block = working_metadata_file$Ind,
  correlation = all_dge_cor_fit$consensus.correlation
)

# head(all_dge_design)

# Defining contrasts
all_dge_contrast_matrix <- makeContrasts(
#Setting my reference control
    Human_vs_Chimp_H2O =
    SpeciesH,
# -------------------------------------------------------------------------    

# DMSO comparisons 
  Chimp_DMSO_vs_H2O =
    StimulusDMSO,
  Human_DMSO_vs_H2O =
    StimulusDMSO + SpeciesH_StimulusDMSO,
  Human_vs_Chimp_DMSO =
    SpeciesH + SpeciesH_StimulusDMSO,
  Interaction_DMSO = 
    SpeciesH_StimulusDMSO,
# -------------------------------------------------------------------------    

# IMR comparisons    
  Chimp_TNFa_vs_H2O =
    StimulusTNFa,
  Human_TNFa_vs_H2O =
    StimulusTNFa + SpeciesH_StimulusTNFa,
  Human_vs_Chimp_TNFa =
    SpeciesH + SpeciesH_StimulusTNFa,
  TNFa_Interaction =
    SpeciesH_StimulusTNFa,
# -------------------------------------------------------------------------    
  
  Chimp_LPS_vs_H2O =
    StimulusLPS,
  Human_LPS_vs_H2O =
    StimulusLPS + SpeciesH_StimulusLPS,
  Human_vs_Chimp_LPS =
    SpeciesH + SpeciesH_StimulusLPS,
  LPS_Interaction =
    SpeciesH_StimulusLPS,
# -------------------------------------------------------------------------    

# DDR comparisons  
  Human_vs_Chimp_DOX = 
    SpeciesH + SpeciesH_StimulusDOX,
  Chimp_DOX_vs_DMSO =
    StimulusDOX - StimulusDMSO, # This outputs: ((Chimp DOX - Chimp H2O) - (Chimp DMSO - Chimp H2O)) = Chimp DOX - Chimp DMSO
  Human_DOX_vs_DMSO =
    StimulusDOX - StimulusDMSO + # This outputs: Chimp DOX - Chimp DMSO
    SpeciesH_StimulusDOX - # This outputs: (Human DOX vs Human H2O) - (Chimp DOX vs Chimp H2O)
    SpeciesH_StimulusDMSO, # This outputs: -[(Human DMSO vs Human H2O) - (Chimp DMSO vs Chimp H2O)]
# The long equation becomes: Chimp DOX - Chimp DMSO + Human DOX - Human H2O - Chimp DOX + Chimp H2O - Human DMSO + Human H2O + Chimp DMSO - Chimp H2O
# OVERALL, this outputs: (Human DOX - Chimp H2O) - (Human DMSO - Chimp H2O) = **Human DOX - Human DMSO**
# This works because the "vs" symbols between attributes act like minus signs. As a result, terms cancel and we are left with the final result
  DOX_Interaction =
    SpeciesH_StimulusDOX -
    SpeciesH_StimulusDMSO,
# -------------------------------------------------------------------------    
  
  Human_vs_Chimp_NUTL = 
    SpeciesH + SpeciesH_StimulusNUTL,
  Chimp_NUTL_vs_DMSO =
    StimulusNUTL - StimulusDMSO,
  Human_NUTL_vs_DMSO =
    StimulusNUTL - StimulusDMSO +
    SpeciesH_StimulusNUTL -
    SpeciesH_StimulusDMSO,
  NUTL_Interaction =
    SpeciesH_StimulusNUTL -
    SpeciesH_StimulusDMSO,
# -------------------------------------------------------------------------    

# UPR comparisons
  Human_vs_Chimp_TUN = 
    SpeciesH + SpeciesH_StimulusTUN,
  Chimp_TUN_vs_DMSO =
    StimulusTUN - StimulusDMSO,
  Human_TUN_vs_DMSO =
    StimulusTUN - StimulusDMSO +
    SpeciesH_StimulusTUN -
    SpeciesH_StimulusDMSO,
  TUN_Interaction =
    SpeciesH_StimulusTUN -
    SpeciesH_StimulusDMSO,
# -------------------------------------------------------------------------    

  Human_vs_Chimp_THA = 
    SpeciesH + SpeciesH_StimulusTHA,  
  Chimp_THA_vs_DMSO =
    StimulusTHA - StimulusDMSO,
  Human_THA_vs_DMSO =
    StimulusTHA - StimulusDMSO +
    SpeciesH_StimulusTHA -
    SpeciesH_StimulusDMSO,
  THA_Interaction =
    SpeciesH_StimulusTHA -
    SpeciesH_StimulusDMSO,
# -------------------------------------------------------------------------    
  levels = all_dge_design
)

# Applying contrasts across samples
all_dge_fit2 <- contrasts.fit(all_dge_fit, all_dge_contrast_matrix)
all_dge_efit2 <- eBayes(all_dge_fit2)
all_dge_signif_gene_results = decideTests(all_dge_efit2, adjust.method = "BH", p.value = 0.05)

summary(all_dge_signif_gene_results)
length(colnames(all_dge_signif_gene_results))
plotSA(all_dge_efit2, main = "Mean-variance All Response Samples final model trend")
```

## Toptable Assignments
```{r Creating TopTables}
Human_vs_Chimp_H2O <- topTable(all_dge_efit2, coef = "Human_vs_Chimp_H2O", adjust.method = "BH", p.value = 1, number = Inf)
Chimp_DMSO_vs_H2O <- topTable(all_dge_efit2, coef = "Chimp_DMSO_vs_H2O", adjust.method = "BH", p.value = 1, number = Inf)
Human_DMSO_vs_H2O <- topTable(all_dge_efit2, coef = "Human_DMSO_vs_H2O", adjust.method = "BH", p.value = 1, number = Inf)
Human_vs_Chimp_DMSO <- topTable(all_dge_efit2, coef = "Human_vs_Chimp_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
Interaction_DMSO <- topTable(all_dge_efit2, coef = "Interaction_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
Chimp_TNFa_vs_H2O <- topTable(all_dge_efit2, coef = "Chimp_TNFa_vs_H2O", adjust.method = "BH", p.value = 1, number = Inf)
Human_TNFa_vs_H2O <- topTable(all_dge_efit2, coef = "Human_TNFa_vs_H2O", adjust.method = "BH", p.value = 1, number = Inf)
Human_vs_Chimp_TNFa <- topTable(all_dge_efit2, coef = "Human_vs_Chimp_TNFa", adjust.method = "BH", p.value = 1, number = Inf)
TNFa_Interaction <- topTable(all_dge_efit2, coef = "TNFa_Interaction", adjust.method = "BH", p.value = 1, number = Inf)
Chimp_LPS_vs_H2O <- topTable(all_dge_efit2, coef = "Chimp_LPS_vs_H2O", adjust.method = "BH", p.value = 1, number = Inf)
Human_LPS_vs_H2O <- topTable(all_dge_efit2, coef = "Human_LPS_vs_H2O", adjust.method = "BH", p.value = 1, number = Inf)
Human_vs_Chimp_LPS <- topTable(all_dge_efit2, coef = "Human_vs_Chimp_LPS", adjust.method = "BH", p.value = 1, number = Inf)
LPS_Interaction <- topTable(all_dge_efit2, coef = "LPS_Interaction", adjust.method = "BH", p.value = 1, number = Inf)
Human_vs_Chimp_DOX <- topTable(all_dge_efit2, coef = "Human_vs_Chimp_DOX", adjust.method = "BH", p.value = 1, number = Inf)
Chimp_DOX_vs_DMSO <- topTable(all_dge_efit2, coef = "Chimp_DOX_vs_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
Human_DOX_vs_DMSO <- topTable(all_dge_efit2, coef = "Human_DOX_vs_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
DOX_Interaction <- topTable(all_dge_efit2, coef = "DOX_Interaction", adjust.method = "BH", p.value = 1, number = Inf)
Human_vs_Chimp_NUTL <- topTable(all_dge_efit2, coef = "Human_vs_Chimp_NUTL", adjust.method = "BH", p.value = 1, number = Inf)
Chimp_NUTL_vs_DMSO <- topTable(all_dge_efit2, coef = "Chimp_NUTL_vs_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
Human_NUTL_vs_DMSO <- topTable(all_dge_efit2, coef = "Human_NUTL_vs_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
NUTL_Interaction <- topTable(all_dge_efit2, coef = "NUTL_Interaction", adjust.method = "BH", p.value = 1, number = Inf)
Human_vs_Chimp_TUN <- topTable(all_dge_efit2, coef = "Human_vs_Chimp_TUN", adjust.method = "BH", p.value = 1, number = Inf)
Chimp_TUN_vs_DMSO <- topTable(all_dge_efit2, coef = "Chimp_TUN_vs_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
Human_TUN_vs_DMSO <- topTable(all_dge_efit2, coef = "Human_TUN_vs_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
TUN_Interaction <- topTable(all_dge_efit2, coef = "TUN_Interaction", adjust.method = "BH", p.value = 1, number = Inf)
Human_vs_Chimp_THA <- topTable(all_dge_efit2, coef = "Human_vs_Chimp_THA", adjust.method = "BH", p.value = 1, number = Inf)
Chimp_THA_vs_DMSO <- topTable(all_dge_efit2, coef = "Chimp_THA_vs_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
Human_THA_vs_DMSO <- topTable(all_dge_efit2, coef = "Human_THA_vs_DMSO", adjust.method = "BH", p.value = 1, number = Inf)
THA_Interaction <- topTable(all_dge_efit2, coef = "THA_Interaction", adjust.method = "BH", p.value = 1, number = Inf)

save_PHTO_toptable_list <- list(
  "HvC_H2O"=Human_vs_Chimp_H2O,
  "C_DMSOvH2O"=Chimp_DMSO_vs_H2O,
  "H_DMSOvH2O"=Human_DMSO_vs_H2O,
  "HvC_DMSO"=Human_vs_Chimp_DMSO,
  "Int_DMSO"=Interaction_DMSO,
  "C_TNFavH2O"=Chimp_TNFa_vs_H2O,
  "H_TNFavH2O"=Human_TNFa_vs_H2O,
  "HvC_TNFa"=Human_vs_Chimp_TNFa,
  "Int_TNFa"=TNFa_Interaction,
  "C_LPSvH2O"=Chimp_LPS_vs_H2O,
  "H_LPSvH2O"=Human_LPS_vs_H2O,
  "HvC_LPS"=Human_vs_Chimp_LPS,
  "Int_LPS"=LPS_Interaction,
  "HvC_DOX"=Human_vs_Chimp_DOX,
  "C_DOXvDMSO"=Chimp_DOX_vs_DMSO,
  "H_DOXvDMSO"=Human_DOX_vs_DMSO,
  "Int_DOX"=DOX_Interaction,
  "HvC_NUTL"=Human_vs_Chimp_NUTL,
  "C_NUTLvDMSO"=Chimp_NUTL_vs_DMSO,
  "H_NUTLvDMSO"=Human_NUTL_vs_DMSO,
  "Int_NUTL"=NUTL_Interaction,
  "HvC_TUN"=Human_vs_Chimp_TUN,
  "C_TUNvDMSO"=Chimp_TUN_vs_DMSO,
  "H_TUNvDMSO"=Human_TUN_vs_DMSO,
  "Int_TUN"=TUN_Interaction,
  "HvC_THA"=Human_vs_Chimp_THA,
  "C_THAvDMSO"=Chimp_THA_vs_DMSO,
  "H_THAvDMSO"=Human_THA_vs_DMSO,
  "Int_THA"=THA_Interaction
  )

saveRDS(save_PHTO_toptable_list, "data/PHTO_toptable_results.RDS")
```

## Look Up Table for all Ensembls of interest
```{r}
## Extracting all unique Ensembl IDs
PHTO_toptable_results <- readRDS("data/PHTO_toptable_results.RDS")
all_ensembls <- PHTO_toptable_results %>%
  purrr::map(rownames) %>%
  unlist() %>%
  unique()

# length(all_ensembl)
# 
# head(all_ensembl)

## Mapping Ensembls to their respective Symbols and Entrez IDs
PHTO_annotation_df <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = all_ensembls,
  keytype = "ENSEMBL",
  columns = c(
    "ENSEMBL",
    "ENTREZID",
    "SYMBOL"
  )
) %>%
  dplyr::distinct(
    ENSEMBL,
    .keep_all = TRUE
  )


PHTO_lookup_table <- tibble::tibble(
  Ensembl = all_ensembls
) %>%
  dplyr::left_join(
    PHTO_annotation_df,
    by = c("Ensembl" = "ENSEMBL")
  ) %>%
  dplyr::rename(
    Entrez_ID = ENTREZID,
    Symbol = SYMBOL
  )

PHTO_ensembl_to_gene_mapping_sucess <- PHTO_lookup_table %>%
  dplyr::summarise(
    Total = dplyr::n(),
    
    Symbol_mapped = sum(!is.na(Symbol)),
    Symbol_unmapped = sum(is.na(Symbol)),
    
    Entrez_mapped = sum(!is.na(Entrez_ID)),
    Entrez_unmapped = sum(is.na(Entrez_ID))
  )
PHTO_ensembl_to_gene_mapping_sucess

saveRDS(PHTO_lookup_table, "data/PHTO_lookup_table.RDS")
```


## Volcano Plots
```{r Volcano Plots comparing treatment conditions within the same species using limma, fig.height=4, fig.width=6}
PHTO_toptable_results <- readRDS("data/PHTO_toptable_results.RDS")
PHTO_lookup_table <- readRDS("data/PHTO_lookup_table.RDS")
PHTO_toptable_list <- list(
  "HvC_H2O",
  "C_DMSOvH2O",
  "H_DMSOvH2O",
  "HvC_DMSO",
  "Int_DMSO",
  "C_TNFavH2O",
  "H_TNFavH2O",
  "HvC_TNFa",
  "Int_TNFa",
  "C_LPSvH2O",
  "H_LPSvH2O",
  "HvC_LPS",
  "Int_LPS",
  "HvC_DOX",
  "C_DOXvDMSO",
  "H_DOXvDMSO",
  "Int_DOX",
  "HvC_NUTL",
  "C_NUTLvDMSO",
  "H_NUTLvDMSO",
  "Int_NUTL",
  "HvC_TUN",
  "C_TUNvDMSO",
  "H_TUNvDMSO",
  "Int_TUN",
  "HvC_THA",
  "C_THAvDMSO",
  "H_THAvDMSO",
  "Int_THA"
  )

generate_volcano_plot <- function(toptable, list_name) {

  volcano_colors <- toptable %>%
    tibble::rownames_to_column("Genes") %>%
    dplyr::left_join(PHTO_lookup_table %>% 
                tibble::rownames_to_column("Ensemble") %>% 
                dplyr::select(Ensemble, Symbol),
              by = c("Genes" = "Ensemble")) %>% 
    dplyr::mutate(
    id_label = ifelse(
      is.na(Symbol),
      Genes,
      Symbol
    )
  ) %>% 
    dplyr::mutate(
      neglog10AdjP = -log10(pmax(adj.P.Val, 1e-300)),
      color_key = case_when(
        adj.P.Val < 0.05 & logFC > 0  ~ "Up",
        adj.P.Val < 0.05 & logFC < 0 ~ "Down",
        TRUE ~ "NS"
      )
    )

  top_genes <- volcano_colors %>%
  dplyr::arrange(adj.P.Val) %>% 
  dplyr::filter(adj.P.Val < 0.05) %>% 
  slice_head(n = 20)
  
  gene_counts <- volcano_colors %>%
    dplyr::summarise(
      Up = sum(color_key == "Up"),
      Down = sum(color_key == "Down"),
      NS = sum(color_key == "NS")
    )
  
  # genes_of_interest <- volcano_colors %>%
  #   dplyr::filter(id_label %in% c("ASPHD2", "C1R", "PARP12", "GBP2", "TNFSF13B", "TUBB", "GRAMD2A", "FAM107A", "ZCWPW2", "TRIB1", "C1S", "CD47", "SFT2D2", "CIBAR1", "IFITM2", "TESK2", "STAMBPL", "NUP58", "BPGM", "CIMAP3", "CD9", "CFH", "SLC35A4", "ENSG00000284048", "UST", "A2M", "CACHD1", "VPS13C", "FBXO31", "NORAD"))

  p <- ggplot(
    volcano_colors,
    aes(x = logFC, y = neglog10AdjP)
    ) +
    geom_point(
      aes(color = color_key),
      size = 2,
      alpha = 0.6
    ) +
    
   # coord_cartesian(
   #    xlim = c(-12, 12),
   #    ylim = c(0, 20)
   #  ) +
    
    # geom_vline(
    # xintercept = c(-1,1),
    # linetype = "dashed"
    # ) +
    # geom_hline(
    # yintercept = -log10(0.05),
    # linetype = "dashed"
    # ) +
    
    geom_text_repel(
      data = top_genes,
      aes(label = id_label),
      size = 3,
      max.overlaps = Inf,
      box.padding = 0.5
    ) +

    scale_color_manual(
      name = NULL,
      values = c(
        "Up" = "red",
        "Down" = "blue",
        "NS" = "grey50"
      ),
      labels = c(
        "Up" = paste0("Up (n = ", gene_counts$Up, ")"),
        "Down" = paste0("Down (n = ", gene_counts$Down, ")"),
        "NS" = paste0("NS (n = ", gene_counts$NS, ")")
        )
      ) +
      
      labs(
        title = list_name,
        x = "log2FC",
        y = "-log10(Adj.P-value)"
      ) +
    
      theme_classic(
      ) +
      
      theme(
        plot.title = element_text(
          hjust = 0.6,
          size = 16,
          face = "bold"
        ),
        text = element_text(
          size = 12
        )
      ) 
          
      list(
        plot = p,
        top = top_genes,
        counts = gene_counts
        #interesting_genes = genes_of_interest
        )
}

for (i in seq_along(PHTO_toptable_results)) {

  results <- generate_volcano_plot(
    PHTO_toptable_results[[i]],
    PHTO_toptable_list[[i]]
  )
## The code below gives written summaries in the console
  cat(
    "\n",
    PHTO_toptable_list[[i]],
    ": Up =", results$counts$Up,
    "| Down =", results$counts$Down,
    "\n"
  )

  print(results$plot) #This prints the plots
}
```

```{r eval=FALSE, include=FALSE}
ggplot(plot_df, aes(x = Stimulus, y = Count, fill = Status)) +
ggplot(plot_df, aes(x = Stimulus, y = Count, fill = Status)) +
  geom_col(position = "fill") +
  
  geom_text(
    aes(
      label = scales::comma(Count),
      group = Status
    ),
    position = position_fill(vjust = 0.5),
    size = 3
  ) +
  
  scale_fill_manual(
    values = c(
      "H DEG" = "#F4D35E",
      "C DEG" = "#E95D6A",
      "non-DEG" = "grey70"
    )
  ) +
  
  scale_y_continuous(
    labels = scales::percent_format()
  ) +
  
  labs(
    x = NULL,
    y = "Proportion",
    fill = "DEG Status"
  ) +
  
  theme_classic() +
  
  theme(
    axis.text.x = element_text(
      size = 12,
      face = "bold"
    ),
    axis.text.y = element_text(
      size = 12,
      face = "bold"
    ),
    axis.title.y = element_text(
      size = 14,
      face = "bold"
    ),
    legend.title = element_text(
      face = "bold"
    )
  )
```

## Scatter plots
```{r}
plot_species_fc <- function(
    toptable_list,
    human_coef,
    chimp_coef,
    interaction_coef,
    lookup_table,
    treatment,
    control,
    genes_to_label = NULL,
    bold_genes = NULL,
    interaction_fdr = 0.05,
    x_limits = NULL,
    y_limits = NULL
) {
  
  # ---------------------------------------------------------
  # 1. Extract Human logFC
  # ---------------------------------------------------------
  
  human_fc <- toptable_list[[human_coef]] %>%
    tibble::rownames_to_column("Ensembl") %>%
    dplyr::select(Ensembl, logFC) %>%
    dplyr::rename(Human_logFC = logFC)
  
  
  # ---------------------------------------------------------
  # 2. Extract Chimp logFC
  # ---------------------------------------------------------
  
  chimp_fc <- toptable_list[[chimp_coef]] %>%
    tibble::rownames_to_column("Ensembl") %>%
    dplyr::select(Ensembl, logFC) %>%
    dplyr::rename(Chimp_logFC = logFC)
  
  
  # ---------------------------------------------------------
  # 3. Combine Human and Chimp logFC
  # ---------------------------------------------------------
  
  fc_df <- human_fc %>%
    dplyr::inner_join(
      chimp_fc,
      by = "Ensembl"
    ) %>%
    dplyr::left_join(
      lookup_table %>%
        dplyr::select(
          Ensembl,
          Symbol
        ),
      by = "Ensembl"
    ) %>%
    dplyr::mutate(
      Symbol = dplyr::if_else(
        is.na(Symbol),
        Ensembl,
        Symbol
      )
    )
  
  
  # ---------------------------------------------------------
  # 4. Identify significant interaction genes
  # ---------------------------------------------------------
  
  interaction_sig_genes <- toptable_list[[interaction_coef]] %>%
    tibble::rownames_to_column("Ensembl") %>%
    dplyr::filter(adj.P.Val < interaction_fdr) %>%
    dplyr::pull(Ensembl)
  
  
  # ---------------------------------------------------------
  # 5. Determine direction of response
  # ---------------------------------------------------------
  
  interaction_df <- fc_df %>%
    dplyr::filter(
      Ensembl %in% interaction_sig_genes
    ) %>%
    dplyr::mutate(
      Direction = dplyr::case_when(
        Human_logFC > 0 & Chimp_logFC > 0 ~ "Human ↑, Chimp ↑",
        Human_logFC < 0 & Chimp_logFC < 0 ~ "Human ↓, Chimp ↓",
        Human_logFC > 0 & Chimp_logFC < 0 ~ "Human ↑, Chimp ↓",
        Human_logFC < 0 & Chimp_logFC > 0 ~ "Human ↓, Chimp ↑",
        TRUE ~ "Zero/NA"
      )
    )
  
  
  # ---------------------------------------------------------
  # 6. Count interaction directions
  # ---------------------------------------------------------
  
  direction_counts <- interaction_df %>%
    dplyr::count(Direction)
  
  
  # ---------------------------------------------------------
  # 7. Select genes to label
  # ---------------------------------------------------------
  
  if (!is.null(genes_to_label)) {
    
    label_df <- fc_df %>%
      dplyr::filter(
        Symbol %in% genes_to_label
      )
    
  } else {
    
    label_df <- fc_df[0, ]
    
  }
  
  
  # ---------------------------------------------------------
  # 8. Number of significant interaction genes
  # ---------------------------------------------------------
  
  n_interaction <- nrow(interaction_df)
  
  # ---------------------------------------------------------
  # 9. Calculate Pearson correlation and p-value
  # ---------------------------------------------------------
  
  # Remove missing values before correlation test
  cor_df <- fc_df %>%
    dplyr::filter(
      !is.na(Human_logFC),
      !is.na(Chimp_logFC)
    )
  
  # Pearson correlation test
  cor_test <- stats::cor.test(
    cor_df$Human_logFC,
    cor_df$Chimp_logFC,
    method = "pearson"
  )
  
  # Extract Pearson r
  pearson_r <- unname(
    cor_test$estimate
  )
  
  # Extract p-value
  pearson_p <- cor_test$p.value
  
  
  # Format p-value for figure
  pearson_p_label <- ifelse(
    pearson_p < 0.001,
    "p < 0.001",
    paste0(
      "p = ",
      sprintf("%.3f", pearson_p)
    )
  )
  
  
  # Create label for figure
  cor_label <- paste0(
    "Pearson r = ",
    sprintf("%.2f", pearson_r),
    "\n",
    pearson_p_label
  )
  
  # ---------------------------------------------------------
  # 10. Create plot
  # ---------------------------------------------------------
  
  p <- ggplot2::ggplot(
    fc_df,
    ggplot2::aes(
      x = Human_logFC,
      y = Chimp_logFC
    )
  ) +
    
    ggplot2::geom_point(
      color = "grey47",
      alpha = 0.6,
      size = 1
    ) +
    
    ggplot2::geom_point(
      data = interaction_df %>%
        dplyr::filter(Direction != "Zero/NA"),
      ggplot2::aes(color = Direction),
      size = 1.5
    ) +
    
    ggrepel::geom_text_repel(
      data = label_df,
      ggplot2::aes(
        label = Symbol,
        fontface = ifelse(
          Symbol %in% bold_genes,
          "bold.italic",
          "plain"
        )
      ),
      size = 3,
      max.overlaps = Inf,
      box.padding = 2,
      point.padding = 0.15,
      force = 2
    ) +
    
    ggplot2::scale_colour_manual(
      values = c(
        "Human ↑, Chimp ↑" = "maroon",
        "Human ↓, Chimp ↓" = "dodgerblue4",
        "Human ↑, Chimp ↓" = "purple3",
        "Human ↓, Chimp ↑" = "green4"
      )
    ) +
    
    ggplot2::geom_vline(
      xintercept = 0,
      linetype = "dashed"
    ) +
    
    ggplot2::geom_hline(
      yintercept = 0,
      linetype = "dashed"
    ) +
    
    # Pearson r and p-value
    ggplot2::annotate(
      "text",
      x = Inf,
      y = Inf,
      label = cor_label,
      hjust = 1.1,
      vjust = 1.3,
      size = 3
    ) +
    
    ggplot2::theme_classic() +
    
    ggplot2::labs(
      x = paste0(
        "Human ",
        treatment,
        " vs ",
        control,
        " logFC"
      ),
      y = paste0(
        "Chimp ",
        treatment,
        " vs ",
        control,
        " logFC"
      ),
      title = paste0(
        "Cross-species ",
        treatment,
        " response correlation"
      ),
      color = paste0(
        "Interaction Gene\nDirections (n = ",
        n_interaction,
        ")"
      )
    ) +
    
    ggplot2::theme(
      legend.position = c(0.15, 0.80)
    )
  
  # ---------------------------------------------------------
  # 11. Optional axis limits
  # ---------------------------------------------------------
  
  if (!is.null(x_limits) | !is.null(y_limits)) {
    
    p <- p +
      ggplot2::coord_cartesian(
        xlim = x_limits,
        ylim = y_limits
      )
  }
  
  
  # ---------------------------------------------------------
  # 12. Return everything
  # ---------------------------------------------------------
  
  return(
    list(
      plot = p,
      fc_df = fc_df,
      interaction_df = interaction_df,
      direction_counts = direction_counts,
      n_interaction = n_interaction,
      pearson_r = pearson_r,
      pearson_p = pearson_p
    )
  )
}
```

```{r}
comparison_table <- tibble::tribble(
  ~treatment, ~control, ~human_coef,       ~chimp_coef,       ~interaction_coef,
  
  "TNFa",     "H2O",    "H_TNFavH2O",     "C_TNFavH2O",     "Int_TNFa",
  "LPS",      "H2O",    "H_LPSvH2O",      "C_LPSvH2O",      "Int_LPS",
  "DOX",      "DMSO",   "H_DOXvDMSO",     "C_DOXvDMSO",     "Int_DOX",
  "NUTL",     "DMSO",   "H_NUTLvDMSO",    "C_NUTLvDMSO",    "Int_NUTL",
  "TUN",      "DMSO",   "H_TUNvDMSO",     "C_TUNvDMSO",     "Int_TUN",
  "THA",      "DMSO",   "H_THAvDMSO",     "C_THAvDMSO",     "Int_THA"
)

comparison_table
```

```{r}
plot_all_species_fc <- function(
    toptable_list,
    comparison_table,
    lookup_table,
    genes_to_label = NULL,
    bold_genes = NULL,
    interaction_fdr = 0.05,
    x_limits = NULL,
    y_limits = NULL
) {
  
  results <- purrr::pmap(
    
    comparison_table,
    
    function(
      treatment,
      control,
      human_coef,
      chimp_coef,
      interaction_coef
    ) {
      
      plot_species_fc(
        toptable_list = toptable_list,
        human_coef = human_coef,
        chimp_coef = chimp_coef,
        interaction_coef = interaction_coef,
        lookup_table = lookup_table,
        treatment = treatment,
        control = control,
        genes_to_label = genes_to_label,
        bold_genes = bold_genes,
        interaction_fdr = interaction_fdr,
        x_limits = x_limits,
        y_limits = y_limits
      )
      
    }
  )
  
  
  # Name each result according to treatment
  
  names(results) <- comparison_table$treatment
  
  
  return(results)
}
```

```{r}
PHTO_fc_results <- plot_all_species_fc(
  toptable_list = PHTO_toptable_results,
  comparison_table = comparison_table,
  lookup_table = PHTO_lookup_table,
  interaction_fdr = 0.05
)

PHTO_fc_results$TUN$plot

purrr::walk(
  PHTO_fc_results,
  ~ print(.x$plot)
)

interaction_summary <- purrr::imap_dfr(
  PHTO_fc_results,
  function(x, treatment) {
    
    x$direction_counts %>%
      dplyr::mutate(
        Treatment = treatment,
        Total_Interaction_Genes = x$n_interaction
      )
    
  }
) %>%
  dplyr::select(
    Treatment,
    Direction,
    n,
    Total_Interaction_Genes
  )

interaction_summary

PHTO_pearson_summary <- purrr::imap_dfr(
  PHTO_fc_results,
  function(result, treatment) {
    
    tibble::tibble(
      Treatment = treatment,
      Pearson_r = result$pearson_r,
      Pearson_p = result$pearson_p
    )
  }
)

PHTO_pearson_summary
```

## Significant Interaction Gene Column Charts
```{r}
plot_interaction_direction_counts <- function(
    result,
    treatment,
    y_limits = NULL,
    y_breaks = NULL
) {
  
  # ---------------------------------------------------------
  # 1. Count interaction gene directions
  # ---------------------------------------------------------
  
  interaction_counts <- result$interaction_df %>%
    dplyr::filter(Direction != "Zero/NA") %>%
    dplyr::mutate(
      
      Direction_group = dplyr::case_when(
        
        Direction %in% c(
          "Human ↑, Chimp ↑",
          "Human ↓, Chimp ↓"
        ) ~ "Same Direction",
        
        Direction %in% c(
          "Human ↑, Chimp ↓",
          "Human ↓, Chimp ↑"
        ) ~ "Opposite Direction",
        
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::filter(!is.na(Direction_group)) %>%
    dplyr::count(
      Direction_group,
      Direction
    )
  
  
  # ---------------------------------------------------------
  # 2. Calculate total genes in each group
  # ---------------------------------------------------------
  
  x_labels <- interaction_counts %>%
    dplyr::group_by(Direction_group) %>%
    dplyr::summarise(
      total = sum(n),
      .groups = "drop"
    )
  
  
  # ---------------------------------------------------------
  # 3. Create labels
  # ---------------------------------------------------------
  
  same_n <- x_labels %>%
    dplyr::filter(Direction_group == "Same Direction") %>%
    dplyr::pull(total)
  
  opposite_n <- x_labels %>%
    dplyr::filter(Direction_group == "Opposite Direction") %>%
    dplyr::pull(total)
  
  
  # If one category contains zero genes
  if (length(same_n) == 0) {
    same_n <- 0
  }
  
  if (length(opposite_n) == 0) {
    opposite_n <- 0
  }
  
  
  # ---------------------------------------------------------
  # 4. Create plot
  # ---------------------------------------------------------
  
  p <- ggplot2::ggplot(
    interaction_counts,
    ggplot2::aes(
      x = Direction_group,
      y = n,
      fill = Direction
    )
  ) +
    
    ggplot2::geom_col() +
    
    ggplot2::geom_text(
      ggplot2::aes(label = n),
      position = ggplot2::position_stack(vjust = 0.5),
      fontface = "bold",
      color = "white",
      size = 6
    ) +
    
    ggplot2::scale_x_discrete(
      limits = c(
        "Same Direction",
        "Opposite Direction"
      ),
      labels = c(
        
        "Same Direction" = paste0(
          "Same Direction\n(n = ",
          same_n,
          ")"
        ),
        
        "Opposite Direction" = paste0(
          "Opposite Direction\n(n = ",
          opposite_n,
          ")"
        )
      )
    ) +
    
    ggplot2::scale_fill_manual(
      values = c(
        "Human ↑, Chimp ↑" = "maroon",
        "Human ↓, Chimp ↓" = "dodgerblue4",
        "Human ↑, Chimp ↓" = "purple3",
        "Human ↓, Chimp ↑" = "green4"
      ),
      drop = FALSE
    ) +
    
    ggplot2::labs(
      x = NULL,
      y = "Number of DEGs",
      title = paste0(
        treatment,
        " significant interaction genes"
      ),
      fill = NULL
    ) +
    
    ggplot2::coord_flip() +
    
    ggplot2::theme_classic() +
    
    ggplot2::theme(
      
      axis.title.x = ggplot2::element_text(
        size = 24,
        face = "bold"
      ),
      
      axis.title.y = ggplot2::element_text(
        size = 24,
        face = "bold"
      ),
      
      axis.text.x = ggplot2::element_text(
        size = 20,
        face = "bold"
      ),
      
      axis.text.y = ggplot2::element_text(
        size = 20,
        face = "bold"
      ),
      
      plot.title = ggplot2::element_text(
        size = 22,
        face = "bold",
        hjust = 0.5
      )
    )
  
  
  # ---------------------------------------------------------
  # 5. Optional y-axis settings
  # ---------------------------------------------------------
  
  if (!is.null(y_limits) | !is.null(y_breaks)) {
    
    p <- p +
      ggplot2::scale_y_continuous(
        limits = y_limits,
        breaks = y_breaks
      )
  }
  
  
  # ---------------------------------------------------------
  # 6. Return plot AND data
  # ---------------------------------------------------------
  
  return(
    list(
      plot = p,
      interaction_counts = interaction_counts,
      group_totals = x_labels
    )
  )
}
```

```{r, fig.width=12, fig.height=6}
TNFa_direction_result <- plot_interaction_direction_counts(
  result = PHTO_fc_results$TNFa,
  treatment = "TNFα"
)
TNFa_direction_result$plot
TNFa_direction_result$interaction_counts
TNFa_direction_result$group_totals

plot_all_interaction_directions <- function(
    PHTO_fc_results,
    y_limits = NULL,
    y_breaks = NULL
) {
  
  treatment_labels <- c(
    "TNFa" = "TNFα",
    "LPS"  = "LPS",
    "DOX"  = "DOX",
    "NUTL" = "NUTL",
    "TUN"  = "TUN",
    "THA"  = "THA"
  )
  
  
  results <- purrr::imap(
    PHTO_fc_results,
    
    function(result, treatment_name) {
      
      plot_interaction_direction_counts(
        result = result,
        treatment = treatment_labels[[treatment_name]],
        y_limits = y_limits,
        y_breaks = y_breaks
      )
      
    }
  )
  
  
  return(results)
}

PHTO_direction_results <- plot_all_interaction_directions(
  PHTO_fc_results
  # ,
  # y_limits = c(0, 1000),
  # y_breaks = seq(0, 1000, by = 250)
)

PHTO_direction_results$TNFa$plot
PHTO_direction_results$LPS$plot
PHTO_direction_results$TUN$plot
PHTO_direction_results$THA$plot
PHTO_direction_results$NUTL$plot
PHTO_direction_results$DOX$plot


PHTO_direction_summary <- purrr::imap_dfr(
  PHTO_direction_results,
  
  function(result, treatment) {
    
    result$interaction_counts %>%
      dplyr::mutate(
        Treatment = treatment
      )
    
  }
) %>%
  dplyr::select(
    Treatment,
    Direction_group,
    Direction,
    n
  )

PHTO_direction_summary <- PHTO_direction_summary %>%
  dplyr::group_by(Treatment) %>%
  dplyr::mutate(
    Total = sum(n),
    Percent = 100 * n / Total
  ) %>%
  dplyr::ungroup()

PHTO_direction_summary
```

