---
title: "PHTO_RNAseq_Quality_analysis"
author: "Harsha Tamtam"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
#this chuck is to setup defaults for basic settings of the webpage, note that this chunk is R language and the chunk below is CSS language
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	dev = c("png","pdf")
)
```

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```


## Introduction

## PHTO Data

## Library Section
```{r Libraries}
library(tidyverse)
library(dplyr)
library(tibble)
library(scales)
library(cowplot)
library(edgeR)
library(smplot2)
library(ggpubr)
library(workflowr)
library(ComplexHeatmap)
library(RColorBrewer)
library(gridExtra)
library(ggrepel)
library(ggVennDiagram)
library(grid)
library(ggrepel)
library(org.Hs.eg.db)
library(enrichplot)
library(clusterProfiler)
library(circlize)
library(gprofiler2)
library(stringr)
library(readr)
library(GO.db)
library(AnnotationDbi)
```


### File cleaning
```{r}
Emma_metadata <- readRDS("data/Emma_metadata_updated_all.RDS")
Emma_counts <- read.csv("data/Emma_counts_file.csv")

stimulus_cleaned_metadata <- Emma_metadata %>%
  dplyr::filter(!Stimulus %in% c("EtOH", "PFOA", "BPA"))

cleaned_metadata_24hr <- stimulus_cleaned_metadata %>%
  dplyr::filter(Time == 24)

cleaned_counts <- Emma_counts %>% 
  as.data.frame() %>% 
  rename(., Ensembl = gene_id) %>% 
  pivot_longer(-Ensembl, names_to = "Samples", values_to = "raw_counts") %>% 
  left_join(cleaned_metadata_24hr, by = c("Samples" = "Sample_ID")) %>% 
  dplyr::select(Ensembl, raw_counts, Final_sample_name) %>% 
  pivot_wider(names_from = Final_sample_name, values_from = raw_counts) %>% 
  column_to_rownames("Ensembl") %>% 
  dplyr::select(-`NA`)

```

###Emma's sample data distribution
```{r, fig.width=12, fig.height=5}
count_df <- cleaned_counts %>%
  mutate(across(everything(), as.numeric)) %>%
  as.data.frame() %>% 
  rownames_to_column("genes") %>% 
  pivot_longer(-genes, names_to = "Samples", values_to = "count_distributions") %>% 
  mutate(count_distributions = as.numeric(count_distributions))
  
ggplot(count_df, aes(x= Samples, y=count_distributions))+
  geom_boxplot()+
  labs(title = "Individual Count Distributions", y="RNA Count Distributions")+
  theme_bw()+
  theme(plot.title = element_text(size = rel(2), hjust = 0.5),
        axis.title = element_text(size = 15, color = "black"),
        axis.ticks = element_line(linewidth = 1.5),
        axis.text.y = element_text(size =10, color = "black", angle = 0, hjust = 0.8, vjust = 0.5),
        axis.text.x = element_text(size =10, color = "black", angle = 90, hjust = 1, vjust = 0.2),
        strip.text.y = element_text(color = "white"))

ggplot(count_df, aes(x = Samples, y = count_distributions+0.000000000000001)) +
  geom_boxplot()+
  scale_y_log10()+
  labs(title = "Individual Count Distributions",y="RNA Count Distributions")+
  theme_bw()+
  theme(plot.title = element_text(size = rel(2), hjust = 0.5),
        axis.title = element_text(size = 15, color = "black"),
        axis.ticks = element_line(linewidth = 1.5),
        axis.text.y = element_text(size =10, color = "black", angle = 0, hjust = 0.8, vjust = 0.5),
        axis.text.x = element_text(size =10, color = "black", angle = 90, hjust = 1, vjust = 0.2),
        strip.text.y = element_text(color = "white"))
```

### Converting raw counts to raw log2cpm
```{r}

count_matrix <- cleaned_counts %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

log2_cpm <- cpm(count_matrix, log = TRUE) %>% 
  as.data.frame() %>% 
  rownames_to_column("genes") %>%
  as_tibble()

```

### Filtering log2cpm counts
```{r}
gene_means <- log2_cpm %>% 
  column_to_rownames(var = "genes") %>% 
  rowMeans()

PHTO_filtered_counts <- count_matrix[gene_means > 0, ]

PHTO_filtered_log2_cpm <- cpm(PHTO_filtered_counts, log = TRUE)

# saveRDS(PHTO_filtered_log2_cpm,"data/PHTO_filtered_log2_cpm.RDS")

  # this code output the number of rows, i.e: the number of genes where log2 > 0
nrow(log2_cpm)  
nrow(PHTO_filtered_log2_cpm)
  # this code outputs the difference between rows of the df log2_cpm, unfiltered from the rows of the df PHTO_filtered_log2_cpm, which is the number of genes that were removed from filtering, as in the number of genes removed where log2 cpm < 0
nrow(log2_cpm) - nrow(PHTO_filtered_log2_cpm)

# Before filtering
  before_tbl <- log2_cpm %>%
  as.data.frame() %>%
  pivot_longer(-genes, names_to = "Samples", values_to = "log2cpm") %>%
  mutate(Filter = "Before")

# After filtering
  after_tbl <- PHTO_filtered_log2_cpm %>%
  as.data.frame() %>%
  rownames_to_column(var = "genes") %>% 
  pivot_longer(-genes, names_to = "Samples", values_to = "log2cpm") %>%
  mutate(Filter = "After")

# This combines the before and after tables to visualize the original and filtered cpm frequency values 
combined_tbl <- bind_rows(before_tbl, after_tbl)

# This code creates a histogram cpm distribution **before filtering**
ggplot(before_tbl, aes(x = log2cpm, fill = )) +
  geom_histogram(bins = 50, alpha = 0.3, position = ("identity")) +
  theme_bw() +
  labs(title = "Log2 CPM Distributions", x = "log2 CPM", y = "Frequencies before filtering")

# This code creates a histogram cpm distribution **after filtering**
ggplot(after_tbl, aes(x = log2cpm, fill = )) +
  geom_histogram(bins = 50, alpha = 0.3, position = ("identity")) +
  theme_bw() +
  labs(title = "Log2 CPM Distributions", x = "log2 CPM", y = "Frequencies after filtering")

# This code creates a histogram cpm distribution data before and after filtering
ggplot(combined_tbl, aes(x = log2cpm, fill = Filter)) +
  geom_histogram(bins = 50, alpha = 0.5, position = ("identity")) +
    scale_fill_manual(
    values = c(
      "Before" = "lightblue",
      "After" = "darkred")
    ) +
  theme_bw() +
  labs(title = "Log2 CPM Distributions", x = "log2 CPM", y = "Frequency")

# This code creates box plots cpm distribution data before and after filtering
reordered_combined_tbl <- combined_tbl %>%
  mutate(Filter = factor(Filter, levels = c("Before", "After"))) 

ggplot(reordered_combined_tbl, aes(x = Filter, y = log2cpm, fill = Filter)) +
  geom_boxplot(outlier.shape = NA) +
  theme_bw() +
  labs(title = "Log2 CPM Boxplots",y = "log2 CPM")

```


### Observing how samples compare with one another

```{r Scatter plot to compare sample similarities under species and treatment categories}
scatter_function <- function(log2_filtered_df, sample_name){
  
after_filter_with_metadata <- log2_filtered_df %>%
  left_join(cleaned_metadata_24hr, by = c("Samples" = "Final_sample_name")) %>%  
  dplyr::select(genes, log2cpm, Samples) %>% 
  dplyr::filter(Samples %in% c("H2O_24_C2", sample_name))

wide_scatter <- after_filter_with_metadata %>%
  pivot_wider(names_from = Samples, values_from = log2cpm) %>% 
  ggplot(., aes(x = H2O_24_C2, y = .data[[sample_name]])) +
  geom_point() +
  stat_cor(method = "pearson") +
  theme_classic()
}
```


```{r This code calls the scatter_function from above and applies a list that I have provided below, fig.height=5, fig.width=7}
lapply(X = c("H2O_24_C1", "DMSO_24_C2", "TUN_24_C2", "THA_24_C2", "DOX_24_C2", "NUTL_24_C2", "LPS_24_C2", "TNFa_24_C2", "H2O_24_H2"), FUN = function(x) {scatter_function(after_tbl, x)})
```


### Heatmap of spearman correlations
```{r, fig.width=13, fig.height=11}
spearman_df <- after_tbl %>% 
  left_join(cleaned_metadata_24hr, by = c("Samples" = "Final_sample_name")) %>% 
  select(genes, log2cpm, Samples) %>%
  pivot_wider(names_from = Samples, values_from = log2cpm) %>% 
  column_to_rownames(var = "genes")

# saveRDS(spearman_df,"data/Ensembl_Condition_filtered_log2cpm_counts_df.RDS")

#head(spearman_df)

attribute_labels_df <- cleaned_metadata_24hr %>%
   select(Sex, Species, Ind, Stimulus, Final_sample_name) %>%
   column_to_rownames("Final_sample_name") %>% 
   as.data.frame()

#head(attribute_labels_df)
 
spearman_cor <- cor(
   spearman_df,
   method = "spearman"
 )
 
attribute_labels_df <- attribute_labels_df[
  colnames(spearman_cor),
  ,
  drop = FALSE
]
## This is to verify that the spearman_cor data frame and attribute_labels_df have the same ordering so that the heatmap clusters appropriately, we want the output to be TRUE 
# identical(
#   rownames(attribute_labels_df),
#   colnames(spearman_cor)
# )
 
 individual_colors <- structure(
   # brewer.pal(12, "Set3"), ## I would use this line of code if I had 12 unique Individuals. However, I have 14, so I need to use the one below to generate colors outside of the main 12 colors in the brwer.pal structure
   colorRampPalette(brewer.pal(12, "Set3"))(14),
   names = unique(attribute_labels_df$Ind)
 )
 
 stimulus_colors <- structure(
   brewer.pal(8, "Accent"),
   names = unique(attribute_labels_df$Stimulus)
 )
 
 heatmap_anno <- HeatmapAnnotation(
   Species = attribute_labels_df$Species,
   Sex = attribute_labels_df$Sex,
   Stimulus = attribute_labels_df$Stimulus,
   Individual = attribute_labels_df$Ind,
   col = list(
     Species = c(
       C = "#F1B72B",
       H = "#8B006D"
        ),
     Sex = c(
       M = "#3C042F",
       F = "#DF707E"
        ),
     Stimulus = stimulus_colors
     ,
     Individual = individual_colors
     ),
   annotation_name_gp = grid::gpar(
    fontsize = 8,
    fontface = "bold"
  )
)
 
 Heatmap(
   spearman_cor,
   name = "Correlation Gradient",
   width = unit(24, "cm"),
   height = unit(20, "cm"),
   row_names_gp = gpar(fontsize = 6),
   column_names_gp = gpar(fontsize = 6),
   top_annotation = heatmap_anno
 )

```


### Scree plot
```{r}
pca_df <- t(spearman_df)
pca_plotting <- prcomp(pca_df,
                       center = TRUE,
                       scale. = FALSE,
                       retx = TRUE
                     )


pca_df <- as.data.frame(pca_plotting$x)

head(pca_df)

pca_variance <- (
  pca_plotting$sdev^2 /
  sum(pca_plotting$sdev^2)
  )*100

pca_analysis <- data.frame(
  Component = 1:length(pca_plotting$sdev),
  Variance = pca_variance
)

head(pca_analysis)

pca_analysis$Samples <- rownames(pca_df)

pca_anno <- pca_df %>%
  rownames_to_column("Samples") %>% 
  left_join(cleaned_metadata_24hr,
    by = c("Samples" = "Final_sample_name")
    )

head(pca_anno)

ggplot(
  pca_analysis %>%
    dplyr::filter(Component <= 10),
  aes(x = Component, y = Variance)
  ) +
  geom_bar(
    stat = "identity",
    fill = "maroon"
  ) +
  geom_line(
    aes(group = 1)
  ) +
  geom_point() +
  scale_x_continuous(
    breaks = 1:10
  ) +
  xlab("Principal Components") +
  ylab("Proportion of Variance (%)") +
  ggtitle("Analysis of Components Contributing to Variance") +
  theme_bw()
```


### Observations on which factors have the most significant effects on principal component variances

```{r This chunk of code is my attempt to make a function that will plot the PCA plots I am interested with in one go}

pca_plot_new <- function(df, pc_x = 1, pc_y = 2, col_var, shape_var, pca_plotting, title = "") {
  
  # variance explained
  pca_variance <- (pca_plotting$sdev^2 / sum(pca_plotting$sdev^2))*100
  
  # axis labels
  x_lab <- sprintf("PC%d (%.1f%%)", pc_x, pca_variance[pc_x] * 100)
  y_lab <- sprintf("PC%d (%.1f%%)", pc_y, pca_variance[pc_y] * 100)
  
  ggplot(df, aes(x = .data[[paste0("PC", pc_x)]], 
                 y = .data[[paste0("PC", pc_y)]], 
                 color = !!sym(col_var), 
                 shape = !!sym(shape_var))) +
    geom_point(size = 5) +
    labs(title = title, x = x_lab, y = y_lab) +
    scale_color_manual(values = c(
      "#8B006D","#DF707E","#F1B72B",
      "#3386DD","#707031","#41B333"
    ))
}

pca_plot <-
  function(pca_df, col_var = NULL, shape_var = NULL, title = "") {
    ggplot(df) + geom_point(aes(
      x = PC1,
      y = PC2,
      color = col_var,
      shape = shape_var
    ),
    size = 5) +
      labs(title = title, x = "PC 1", y = "PC 2") +
      scale_color_manual(values = c(
        "#8B006D",
        "#DF707E",
        "#F1B72B",
        "#3386DD",
        "#707031",
        "#41B333"
      ))
  }

get_regr_pval <- function(model) {
  # Returns the p-value for the F statistic of a linear model
  # mod: class lm
  stopifnot(inherits(model, "lm"))
  fstat <- summary(model)$fstatistic
  pval <- 1 - pf(fstat[1], fstat[2], fstat[3])
  return(pval)
}

plot_versus_pc <- function(pca_anno, pc_num, facs) {
  # df: data.frame
  # pc_num: numeric, specific PC for plotting
  # facs: column name of df for plotting against PC
  pc_char <- paste0("PC", pc_num)
  # Calculate F-statistic p-value for linear model
  
  form <- as.formula(paste(pc_char, "~", facs))
  model <- lm(form, data = pca_anno)

  pval <- get_regr_pval(model)

  if (is.numeric(pca_anno[[facs]])) {
    ggplot(pca_anno, aes_string(x = facs, y = pc_char)) +
      geom_point() +
      geom_smooth(method = "lm") +
      labs(title = sprintf("p-val: %.3g", pval))
  }
  else {
    ggplot(pca_anno, aes_string(x = facs, y = pc_char)) +
      geom_boxplot() +
      labs(title = sprintf("p-val: %.3g", pval))
  }
}

#The lines of code below ouput the number of unique labels for each attribute I call from the pca_charts data frame 
# sapply(c("species", "tx", "ind", "sex"), function(x) {
#   length(unique(na.omit(pca_charts[[x]])))
# })

```


```{r Here I will apply the function about to this current function with a list to plot all PCA variance graphs of interest, fig.height=8, fig.width=7}
rownames(pca_anno) <- pca_anno$Samples

pca_converted <- pca_anno %>%
  tidyr::separate(
    Samples,
    into = c("stimulus", "time", "individual"),
    sep = "_"
  ) %>%
  mutate(
    species = case_when(
      startsWith(individual, "H") ~ "Human",
      startsWith(individual, "C") ~ "Chimp",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::select(
    species,
    stimulus,
    individual,
    everything()
  )

pca_charts <- pca_converted %>% 
  mutate(species=factor(species, levels = c("Chimp", "Human"))) %>% 
  mutate(stimulus=factor(stimulus, levels = c("H2O", "DMSO", "DOX", "NUTL", "LPS", "TNFa", "TUN", "THA"))) %>% 
  mutate(sex=factor(Sex, levels=c("F","M")))

facs <- c("species", "sex", "stimulus", "individual")
names(facs) <- c("Species", "Sex", "Stimulus", "Individual")

plots_of_pcas <- lapply(names(facs), function(f_label) {
  f <- facs[f_label]
  f_v_pc1 <- plot_versus_pc(pca_charts, 1, f)
  f_v_pc2 <- plot_versus_pc(pca_charts, 2, f)
  f_v_pc3 <- plot_versus_pc(pca_charts, 3, f)
  f_v_pc4 <- plot_versus_pc(pca_charts, 4, f)
  f_v_pc5 <- plot_versus_pc(pca_charts, 5, f)
  f_v_pc6 <- plot_versus_pc(pca_charts, 6, f)

  grid.arrange(f_v_pc1, f_v_pc2, f_v_pc3, f_v_pc4, f_v_pc5, f_v_pc6, ncol = 2, top = paste0(f_label," PCA Analyses"))
})

```

### PCA plot demonstrating Species/Sex/Individual (PC1) and Treatment (PC2) stratification
```{r This chunk of code is my attempt at creating PCA plots to illustrate which factors contribute to varience between samples}

stimulus_shape <- scale_shape_manual(
  values = c("DMSO" = 16,
             "DOX" = 17,
             "H2O" = 18,
             "LPS" = 19,
             "NUTL" = 20,
             "THA" = 21,
             "TNFa" = 22,
             "TUN" = 23
             ),
  name = "Stimuli"
)

ggplot(
  pca_converted,
  aes(
    x = PC1,
    y = PC2,
    color = Species,
    shape = stimulus
  )
) +
  geom_point(size = 4) +
  geom_text_repel(
    aes(label = rownames(pca_converted)), size = 2.5) +
  stimulus_shape +
  labs(
    title = "PCA of RNA-seq Samples",
    x = paste0(
      "PC1: ",
      round(pca_variance[1], 1),
      "% variance"
    ),
    y = paste0(
      "PC2: ",
      round(pca_variance[2], 1),
      "% variance"
    )
  ) +
  theme_classic()

```

